 

### 2021-06-09 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests. Dense canopy forests composed of broadleaf trees can have high $LAD$ in the upper canopy layers, as well as in the understory for eg. in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a]. In open canopy forests, such as needle-leaf forests, LAD is predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (@conditQuantifyingDeciduousnessTropical2000c, Meakem, *temperate REFS*). In this review, we focus on growing season conditions unless otherwise noted.

@@

**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
The $T_{air}$ gradient along canopies and understories vary with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. Noteably, this pattern can be reversed in open canopy forests or sparse stands, particularly in xeric sites, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the canopy and the soil that is intercepted by lower-canopy layers--analogous to 'canopy greenhouse effect' [@banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].

@@

### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
Leaf traits vary within an indivual tree at a vertical gradient from exposed canopy regions to shaded regions (intraspecific variation), and as well across mutiple trees' exposed canopy to shaded canopy regions (inter-specific variation).

@@

Exposed-canopy leaves are better adapted to heat and water stress than shaded-leaves where sun-leaves are often thicker, smaller, hairier, with greater leaf dissections [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], hydraulically efficient [Brodribb et al. 2010], with structural strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*]. All, collectively regulate $T_{leaf}$-photosynthetic balance. Whereas, shaded leaves have lower convective cooling strategies, and invest more in light capture [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al. 2019]. Under homogeneous light conditions, sun-leaf differences in canopy orientations are shaped by canopy $T_{air_{max}}$ and $VPD$ stress than solely by light [@mediavillaFoliarPlasticityRelated2019a].

@@

The distribution of biochemical traits enable photosynthesis and metabolic processes, and protect against heat and light damage in sun-leaves. $N_{area}$, that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition along sun-leaves and, shade-leaves during sunflecks [@kitaoCanopyNitrogenDistribution2018]. Similarly, carotenoids and xanthophyll pigments-violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function- convert excess light to heat dissaptation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. Constraints to biochemical traits, particularly $VAZ$ can curb increasing temperature acclimation [@kumarathungeAcclimationAdaptationComponents2019a]. Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed-canopies, making these species particularly advantageous [*REF*].

@@

Widely, across forest species, fundamental trait adaptation to light gradients are similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], that further vary with levels of trait plasticity [REF], for eg. between  canopy multi-species, and understory species, trait plasticity can vary several-folds based on inherent specie-specific traits, even under similar biophysical conditions respectively. Spatio-temporal gradients, offer another axis for trait variation, as they are heterogenous in time and space, with different levels of plasticity [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a]. Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000]. Therefore $T_{leaf}$ varies across specie-specific traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase [@fauset_differences_2018; @songTreeSurfaceTemperature2020].

@@

Deciduousness as a trait is an adaptation to converse water under stress. Indeed, greater proportion of deciduous species are found in drier tropical regions than wetter. Among tropical species individuals that can be deciduous, greater proportion are in larger size classes than understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000c]. In deciduous-temperate forests, spatio-temporal leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al. 2013; @seiwaChangesLeafPhenology1999].Therefore, deciduous leaf-loss or -gain can alter biophysical and $T_{leaf}$ gradient [@rey-sanchez_spatial_2016].

@@

Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] where although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006]. This diurnal pattern and resulting $A_{max}$ differs between species, such as early- and late-successional [@ishiiExploringRelationshipsCanopy2004a], and between gymnosperms and angiosperms [@bonan_ecological_2016].

>>>>

LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests. Dense canopy forests composed of broadleaf trees can have high $LAD$ in the upper canopy layers, as well as in the understory such as in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs. bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a]. In open canopy forests, such as needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs. open canopy forests [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016]. In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (@conditQuantifyingDeciduousnessTropical2000c, Meakem, *temperate REFS*). In this review, we focus on growing season conditions unless otherwise noted.

@@

**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
The $T_{air}$ gradient along canopies and understories vary with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. Noteably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers--analogous to 'canopy greenhouse effect' [@banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].

@@

### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
Leaf trait comparison within an individual tree here is referred to as intraspecific variation, and comparison of traits between species of multiple trees here is referred to as inter-specific variation.

@@

Exposed-canopy leaves are better adapted to heat and water stress than shaded-leaves where sun-leaves are often thicker, smaller, hairier, with greater leaf dissections [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], hydraulically efficient [Brodribb et al. 2010], with structural strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*]. All, collectively regulate $T_{leaf}$-photosynthetic balance. Whereas, shaded leaves have lower convective cooling strategies, and invest more in light capture [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al. 2019]. Under homogeneous canopy light conditions, sun-leaf differences in canopy orientations are shaped by canopy $T_{air_{max}}$ and $VPD$ stress than solely by light [@mediavillaFoliarPlasticityRelated2019a].

@@

The vertical distribution of biochemical traits enables photosynthesis and metabolic processes, and protects against heat and light damage in exposed canopy leaves. Particualy, $N_{area}$ that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition along sun-leaves and, shade-leaves during sunflecks [@kitaoCanopyNitrogenDistribution2018]. Similarly, carotenoids and xanthophyll pigments-violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function- convert excess light to heat dissaptation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. Constraints to biochemical traits, particularly $VAZ$ can curb increasing temperature acclimation [@kumarathungeAcclimationAdaptationComponents2019a]. Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed-canopies, making these species particularly advantageous [*REF*].

@@

Widely, across forest species, the pattern of fundamental trait adaptation to light gradients is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. However, trait plasticity can vary several-folds between  multi-species canopy and understory,  based on their inherent specie-specific traits and environmental conditions. Spatio-temporal gradients, offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a]. Along the ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ are observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000]. Therefore, collectively, $T_{leaf}$ differs across specie-specific traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].

@@

Deciduousness as a trait is an adaptation to conserve water under stress. Indeed, greater proportion of deciduous species are found in drier tropical regions than wetter. Among tropical species individuals that can be deciduous, greater proportion are in larger size classes than understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000c]. In deciduous-temperate forests, spatio-temporal leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al. 2013; @seiwaChangesLeafPhenology1999].Therefore, deciduous leaf-loss or -gain can alter biophysical and $T_{leaf}$ gradient [@rey-sanchez_spatial_2016].

@@

Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] where although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steep in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006]. This diurnal pattern and resulting $A_{max}$ varies between species, such as early- and late-successional [@ishiiExploringRelationshipsCanopy2004a], and between gymnosperms and angiosperms [@bonan_ecological_2016].
